Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-25 survey/survey.tex
some general cleanup

\title{Embracing the Laws of Physics: \\ Reversible Models of Computation}
\author{Jacques Carette, Roshan P. James, Amr Sabry}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction : Reversibility, the Missing Principle}

@@

Traditional models of computation are all unityped, generally
operating on unstructured sequences of symbols. But recent
investigations, in the intersection between mathematics and
computer science (particularly in Homotopy Type Theory~\cite{hott-book})
reveal that not only can mathematics be typed, but that doing so
reveals further structure -- in both mathematics and in computation.

@@

\subsection{From classical to quantum}

@@

\subsection{conservation of information}

@@

This stands in stark contrast with most current approacheds to
reversible computation, which start from current models of
computation involving irreversible operations and try to find
various ways to \emph{patch things up} so as to be reversible.

@@

Starting from a physical perspective, whatever our notion of data
is, we will be interested in programs as representating transformations
of that data which are reversible. In other words, we want our
programs-as-transformations to ``play well'' with the inherent notion
of ``sameness'' that our data will carry. Thus we need to start by
looking at what structure our data has, and that will help us define
an appropriate notion of program. Of course, when programs themselves
are data, things do get more complicated.  In the following sections,
we will look at different natural classes of data, and explore the
corresponding notion of reversible programs.

@@

Our first instinct would be to similarly replace $=$ with
$\equiv$, i.e. propositional equality. This bring the second
problem to the fore: this is false. For example, unless $A \equiv B$,
$A \times B \equiv B \times A$ is not normally provable%
\footnote{Not without assuming that $U$ is univalent, which we
don't.}. But it should be clear that $A \times B$ and $B \times A$
contain equivalent information. In other words, we would like to
be able to witness that $A \times B$ can be reversibly deformed
into $B \times A$. Which motivates the introduction of 
type \emph{equivalences}. To do this, we need a few important
auxilliary concepts first.

@@

\noindent Why \emph{quasi}? The reasons are beyond our scope, but
the interested reader can read sections $2.4$ and $4$ in \cite{hott-book}.
There are several conceptually different, but equivalent, 
``better'' definitions.  We record just one here:

@@

We now have in our hands our desired denotational semantics: we want
to create a language $\Pi$ such that the types and type combinators
map to $\bot, \top, \uplus, \times$, and such that we have ground
terms whose denotation are all $16$ type isormorphisms of
Figure~\ref{type-isos}. This is completely straightforward, as
we can simply do this literally. To make the analogy with 
commutative semirings stand out even more, we will use $0, 1, +$, and $*$
at the type level, and will denote ``equivalence'' by $\iso$.
Thus Figure~\ref{pi-terms} shows the ``constants'' of the language.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data IV: Recursive Types}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data V: Exclusive Disjunctions}

???

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements} many students and colleagues

\bibliographystyle{plainnat}
%\pagestyle{headings}
\bibliography{cites}
\end{document}

>>>>

\title{Embracing the Laws of Physics: \\ Three Reversible Models of Computation}
\author{Jacques Carette \qquad\qquad Roshan P. James \qquad\qquad Amr Sabry \\
McMaster University \qquad\qquad Google \qquad\qquad Indiana University}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reversibility, the Missing Principle}

\subsection{Maxwell's Demon}

@@

Traditional models of computation are all unityped, generally
operating on unstructured sequences of symbols. But recent
investigations, in the intersection between mathematics and
computer science (particularly in Homotopy Type Theory~\cite{hottbook})
reveal that not only can mathematics be typed, but that doing so
reveals further structure -- in both mathematics and in computation.

@@

\subsection{From Classical to Quantum}

@@

\subsection{Conservation of Information}

@@

This stands in stark contrast with most current approaches to
reversible computation, which start from current models of
computation involving irreversible operations and try to find
various ways to \emph{patch things up} so as to be reversible.

@@

Starting from a physical perspective, whatever our notion of data
is, we will be interested in programs as representing transformations
of that data which are reversible. In other words, we want our
programs-as-transformations to ``play well'' with the inherent notion
of ``sameness'' that our data will carry. Thus we need to start by
looking at what structure our data has, and that will help us define
an appropriate notion of program. Of course, when programs themselves
are data, things do get more complicated.  In the following sections,
we will look at different natural classes of data, and explore the
corresponding notion of reversible programs.

@@

Our first instinct would be to similarly replace $=$ with
$\equiv$, i.e. propositional equality. This bring the second
problem to the fore: this is false. For example, unless $A \equiv B$,
$A \times B \equiv B \times A$ is not normally provable%
\footnote{Not without assuming that $U$ is univalent, which we
don't.}. But it should be clear that $A \times B$ and $B \times A$
contain equivalent information. In other words, we would like to
be able to witness that $A \times B$ can be reversibly deformed
into $B \times A$. Which motivates the introduction of 
type \emph{equivalences}. To do this, we need a few important
auxiliary concepts first.

@@

\noindent Why \emph{quasi}? The reasons are beyond our scope, but
the interested reader can read sections $2.4$ and $4$ in \cite{hottbook}.
There are several conceptually different, but equivalent, 
``better'' definitions.  We record just one here:

@@

We now have in our hands our desired denotational semantics: we want
to create a language $\Pi$ such that the types and type combinators
map to $\bot, \top, \uplus, \times$, and such that we have ground
terms whose denotation are all $16$ type isomorphisms of
Figure~\ref{type-isos}. This is completely straightforward, as
we can simply do this literally. To make the analogy with 
commutative semirings stand out even more, we will use $0, 1, +$, and $*$
at the type level, and will denote ``equivalence'' by $\iso$.
Thus Figure~\ref{pi-terms} shows the ``constants'' of the language.

@@

@@

Data IV: Recursive Types

Data V: Exclusive Disjunctions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements} We would like to thank the numerous
students and colleagues who participated in various aspects of this
research and who provided valuable feedack and constructive criticism.

\bibliographystyle{acm}
%\pagestyle{headings}
\bibliography{cites}
\end{document}
